---
title: "The RNA Centric Analysis System Report"
author: "Bora Uyar, Dilmurat Yusuf, Ricardo Wurmus, Altuna Akalin"
date: "`r Sys.time()`"
params: 
  dbPath: 'RCAS.sqlite'
  workdir: '.'
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 4.5, fig.width = 8)
knitr::opts_knit$set(progress = FALSE, root.dir = params$workdir)
```

```{r setup}
library(RCAS)
library(DBI)
library(ggplot2)
library(pheatmap)
mydb <- RSQLite::dbConnect(RSQLite::SQLite(), params$dbPath)
```
# How similar are these samples based on the gene overlap matrix?

```{r gene_overlaps}
geneOverlaps <- dbReadTable(mydb, 'geneOverlaps')

SM <- as.matrix(proxy::dist(x = as.matrix(geneOverlaps[,c(-1)]), method = 'Jaccard', by_rows = FALSE, diag = TRUE, upper = TRUE))
pheatmap(SM, display_numbers = FALSE)
```


# how similar are these samples based on their distribution at different transcript features? 
```{r annotation_summaries}
AS <- dbReadTable(mydb, 'annotationSummaries')
annotationSummaries <- as.matrix(AS[,c(-1)])
rownames(annotationSummaries) <- AS$sampleName
pheatmap(annotationSummaries, scale = 'row')
```


```{r coverage_profiles}
cvg <- dbReadTable(mydb, 'featureBoundaryCoverageProfiles')
cvg$sampleGroup <- stringi::stri_extract(cvg$sample, regex = '^.{4}')
p <- ggplot(cvg, aes(x = bases, y = meanCoverage)) + geom_line(aes(colour = sampleGroup), fill = FALSE) + facet_grid(feature ~ boundary) + geom_vline(xintercept = 0)
plotly::ggplotly(p)
```

# Acknowledgements

RCAS is developed in the group of 
[Altuna Akalin](http://bioinformatics.mdc-berlin.de/team.html#altuna-akalin-phd)
(head of the Scientific Bioinformatics Platform) by
[Bora Uyar](http://bioinformatics.mdc-berlin.de/team.html#bora-uyar-phd)
(Bioinformatics Scientist),
[Dilmurat Yusuf](http://bioinformatics.mdc-berlin.de/team.html#dilmurat-yusuf-phd)
(Bioinformatics Scientist) and 
[Ricardo Wurmus](http://bioinformatics.mdc-berlin.de/team.html#ricardo-wurmus)
(System Administrator) at the Berlin Institute of Medical Systems Biology
([BIMSB](https://www.mdc-berlin.de/13800178/en/bimsb))
at the Max-Delbrueck-Center for Molecular Medicine
([MDC](https://www.mdc-berlin.de)) in Berlin.

RCAS is developed as a bioinformatics service as part of
the [RNA Bioinformatics Center](http://www.denbi.de/index.php/rbc),
which is one of the eight centers of
the German Network for Bioinformatics Infrastructure
([de.NBI](http://www.denbi.de/)).  

# Session Information
```{r sessionInfo}
sessionInfo()
```
